The Name of the Title Is Hope

### 2025-01-17 paper.tex
Rewrites.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers. To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory. Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans is memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}. Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization. In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like \emph{scan} can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies. However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and status flags to ensure coherent views of data. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and incorporate the reductions of all preceding tiles. In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the progress of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

@@

\subsubsection{Chained-Scan, Decoupled-Lookback}
Building on \emph{Stream-Scan}, Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on progress guarantees, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Ideal Scan Architecture}

@@

Additionally, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.

As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}. This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and certain ARM GPUs~\cite{sorensen 2}. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is on exploring and formalizing progress models and their implications for synchronization in GPU programming frameworks.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE.}

In \emph{Chained-Scan}, workgroups operate in a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, while subsequent workgroups consume the reductions of all preceding workgroups to complete the scan. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than static assignment based on virtualized workgroup indices. While this dynamic assignment guarantees that predecessor work tiles are either scheduled or completed, without FPG there is no guarantee that these reductions will eventually become available. A scheduler lacking FPG may fail to allocate sufficient execution time to certain workgroups, leaving their reductions incomplete indefinitely. In such cases, consuming workgroups stall during their lookback phase, waiting for reductions that are never posted to global memory.

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. \john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} \thomas{I actually do not know the answer to this question. My impression is that (lack of) FPG is purely a phenomenon of the subgroup scheduler, and thus pigeonholing a multiprocessor with a single workgroup \emph{would} solve the problem. Of course this would be totally infeasible to implement due to varied multiprocessor resources. HOWEVER, do workgroups execute \emph{completely} indepently of one another?} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence across all hardware is a critical issue for developers. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. In \emph{ScalarFallback}, each time a workgroup spins on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.

@@

\section{Decoupled-Fallback}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}
Our \emph{Decoupled-Fallback} method is a further extension of the \emph{Chained-Scan} architecture.

@@

\subsection{Split Method}
\begin{enumerate}
    \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
    \item Pro: Full 32 or 64 bits.
    \item Pro: No memory model or 64 bit atomics required!
    \item Pro: Can safely fallback insert without risking clobbering.
    \item Con: 4x Memory usage
    \item Con: Tad slower, more complicated.
\end{enumerate}

@@

\subsubsection{Size Agnostic Scan}
For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.

The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.

>>>>

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers. To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory. Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans are memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}. Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization. In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization.

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies. However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and incorporate the reductions of all preceding tiles. In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

@@

\subsubsection{Chained-Scan, Decoupled-Lookback}
Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Ideal Scan Architecture}
\thomas{This will be rewritten. It is clear to me now that the focus should be on the tradeoffs between the Oversubcription and not-Oversubcription(Persistent-Thread or Fixed Launch) paradigms. We want to communicate to the reader that because we want portability, we want to remain within the Oversubcription paradigm. We LIKE Oversubcription. We want to reframe the downsides as "things that happen when you step outside of Oversubcription." I.e. when you step outside of Oversubcription, now YOU have to manage occupancy -> occupancy discovery. When you step outside of Oversubcription now YOU have to worry about launch-related global memory access paterns -> overpartitioning your rake. Then we want to say although Oversubcription precludes inter-workgroup barriers besides kernel launches, thats OK because Chained Scan does not require an explicit barrier. Thus, Chained-Scan is good \emph{because} it lets us stay Oversubscribed}.

@@

Additionally, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

As a consequence of its single kernel execution and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks required additional kernel launches and bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, enabling its application to a variety of practical routines, including data deduplication, sparse data compaction, error correction, and interval merging.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}. This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen 2}. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
  \begin{enumerate}
    \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
    \item Circular Ordering: A circular chain of tasks exists.
  \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1-3 are easily satisfied due to the one-to-one mapping of work-tiles to workgroups. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled independently \thomas{Possibly confusing. . . Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction. This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.

A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm. In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE.}

\thomas{This example may be redundant now.}
For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence on key segments of hardware is a critical issue for developers. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets the mutual exclusion condition outlined by Coffman et al., ensuring that no single workgroup holds exclusive control over a resource. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.

@@

\section{Design}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}

\subsection{Breaking Deadlocks}
\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing  Coffman's mutual exclusion is the best way to do it, under our given constraints.}

Deadlock in \emph{Chained-Scan} arises from the inter-workgroup dependencies that are inherent to scan operations. These dependencies are governed by Coffman et al.'s~\cite{10.1145/356586.356588} four necessary conditions for deadlock: \emph{mutual exclusion}, \emph{resource holding}, \emph{no preemption}, and \emph{circular dependency}. While all four conditions are theoretically avenues for addressing deadlock, practical constraints and the requirements of \emph{Chained-Scan} significantly narrow the set of viable approaches.

The first condition, \emph{dependency relationships}, is immutable in the context of scan operations. Similarly, \emph{no preemption} is immutable within the GPU programming model. Addressing \emph{resource holding} by allowing workgroups to release resources (losing context) during waits effectively reintroduces kernel launches as inter-workgroup barriers. Designing tile sizes to break \emph{circular dependency} by mapping workgroups one-to-one with multiprocessors might seem feasible but is impractical. GPU vendors provide no guarantees regarding workgroup progress behavior, leaving starvation risks unresolved. Moreover, such strategies are non-portable and contradict the goal of supporting diverse hardware architectures.

\subsection{Decoupled-Fallback}

@@

\subsection{Split Method}
\begin{enumerate}
    \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
    \item Pro: Full 32 or theoretically 64 bits.
    \item Pro: No memory model or 64 bit atomics required!
    \item Pro: Can safely fallback insert without risking clobbering.
    \item Con: 2x Memory usage
    \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
\end{enumerate}

@@

\subsubsection{Size Agnostic Scan}
For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network. The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
